---
title: "Introduction to ggplot2"
subtitle: "(and a little tidyverse)"
author: "Jacob Spangler"
date: "2019-05-18"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

<style type="text/css">
.remark-slide-content {
    font-size: 30px;
    padding: 1em 4em 1em 4em;
}
</style>

# Overview

  - Review Plotting in R
  
  - What is ggplot2?
  
  - Examples of ggplot2
  
  - Exploring data
  
  - Conclusions
  
```{r, echo=F, message=F, warning=F}

library(tidyverse)

```

---

# Base R Scatterplot

```{r, fig.height=5, fig.width=5}

plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width)

```

---

# Base R Histogram

```{r, fig.height=5, fig.width=5}

hist(x=iris$Sepal.Length)

```

---

# Base R Boxplots

```{r, fig.height=5, fig.width=5}

boxplot(Sepal.Length ~ Species, data=iris)

```

---

# What is ggplot2?

**ggplot2** is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell **ggplot2** how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.

---

# How to install?

```{r eval=F, tidy=F}

# The easiest way to get ggplot2 is to install the whole tidyverse:
install.packages("tidyverse")

# Alternatively, install just ggplot2:
install.packages("ggplot2")

# Or the the development version from GitHub:
# install.packages("devtools")
devtools::install_github("tidyverse/ggplot2")

```

---

# ggplot2 Scatterplots

```{r, fig.height=5, fig.width=5}

ggplot(iris) +
  aes(x=Sepal.Length, y=Sepal.Width) +
  geom_point()

```

---

# ggplot2 Scatterplots - Diagonal Line

```{r, fig.height=5, fig.width=5}

ggplot(iris) +
  aes(x=Sepal.Length, y=Sepal.Width) +
  geom_point() +
  geom_abline()

```

---

# ggplot2 Scatterplots - Adding Color

```{r, fig.height=5, fig.width=5}

ggplot(iris) +
  aes(x=Sepal.Length, y=Sepal.Width, color=Species) +
  geom_point() +
  geom_abline()

```

---

# ggplot2 Histograms

```{r, fig.height=5, fig.width=5, warning=F, message=F}

ggplot(iris) + 
  aes(x=Sepal.Length) +
  geom_histogram()

```

---

# ggplot2 Histograms - Bin options

```{r, fig.height=5, fig.width=5}

ggplot(iris) +
  aes(x=Sepal.Length) +
  geom_histogram(color="black", binwidth = 0.2)

```

---

# ggplot2 Histograms - Faceting

```{r, fig.height=5, fig.width=5}

ggplot(iris) +
  aes(x=Sepal.Length, fill=Species) +
  geom_histogram(color="black", binwidth = 0.2) +
  facet_wrap(~Species, ncol=1) +
  theme(legend.position = "none")

```

---

# ggplot2 Boxplots

```{r, fig.height=5, fig.width=5, warning=F, message=F}

ggplot(iris) + 
  aes(x=Species, y=Sepal.Length) +
  geom_boxplot()

```

---

# ggplot2 Boxplots - Adding color

```{r, fig.height=5, fig.width=5, warning=F, message=F}

ggplot(iris) + 
  aes(x=Species, y=Sepal.Length, color=Species) +
  geom_boxplot()

```

---

# ggplot2 Boxplots - Adding points

```{r, fig.height=5, fig.width=5, warning=F, message=F}

ggplot(iris) + 
  aes(x=Species, y=Sepal.Length, color=Species) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(height = 0, 
                                        width = 0.3))

```

---

# base R vs ggplot2 - Scatterplots

.pull-left[

```{r, echo=F, fig.height=5, fig.width=5}

plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width)

```

]


.pull-right[

```{r, echo=F, fig.height=5, fig.width=5}

ggplot(iris) +
  aes(x=Sepal.Length, y=Sepal.Width) +
  geom_point()

```

]
---

# DataFest Dataset

```{r, echo=F}

Example_Data <- read_csv("../data/raw/Public Potable Water Systems FINAL 06-22-2018_0.csv",
                         col_types = cols())

```

- What about example data set?
  - There are **`r nrow(Example_Data)`** rows
  - There are **`r ncol(Example_Data)`** columns

---

# DataFest Scatterplot

```{r, fig.height=5, fig.width=7, warning=F, message=F}

ggplot(Example_Data) + 
  aes(x=`Residential Population`, 
      y=`Number of COMBINED Service Connections (CB)`) +
  geom_point(alpha=0.4)

```

---

# DataFest Histogram

```{r, fig.height=5, fig.width=7, warning=F, message=F}

ggplot(Example_Data, aes(x=`Residential Population`)) +
  geom_histogram(color="black")

```

---

# Conclusions

- **ggplot2** has better syntax
- **ggplot2** defaults to prettier graphs
- **ggplot2** has lots of flexibility

```{r, echo=F, fig.height=5, fig.width=6}

cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00",
               "#CC79A7", "#FFFFFF")

ggplot(iris) +
  aes(x=Sepal.Length, y=Sepal.Width, color=Species) +
  geom_point() +
  xlab("Sepal Length") + ylab("Sepal Width") +
  theme(axis.text = element_text(color="black", size=14),
        plot.title = element_text(hjust=0.5, size = 16),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        axis.title = element_text(size=14)) +
  ggtitle("Sepal Length vs Width in Iris data") +
  scale_x_continuous(breaks=c(seq(4, 8, 1)), limits=c(4, 8)) +
  scale_y_continuous(breaks=c(seq(2, 4.5, 0.5)), limits = c(2, 4.5)) +
  scale_color_manual(values=cbPalette)

```


---

# References

- tidyverse: https://www.tidyverse.org/

- R for Data Science: https://r4ds.had.co.nz/index.html

- xaringan: https://github.com/yihui/xaringan
